
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# MB5370: Techniques in Marine Science 1

# Module 2 - Workshop 01. Foundations of Data Science


House Keeping
```{r}
objects()
```


Purge global environment
```{r}
rm(list = ls())
```

Confirm that global session memory is now completely vacant 
```{r}
objects()
```

Install here package
```{r}
# install.packages("here")
```

Load here package
```{r}
library(here)
library(tidyverse)
library(readxl)
```

Practice Import A: Loading a standard comma-separated plain text file
```{r}
benthic_cover <- read_csv("data/reef_cover_log.csv")
```

Practice Import B: Parsing a tab-separated telemetry instrument array string
```{r}
acoustic_stream <- read_tsv("data/acoustic_telemetry_stream.txt")
```

Practice Import C: Targeting a specific sheet in a multi-tab Excel spreadsheet
```{r}
fisheries_annual <- read_excel("data/fish_catch_data.xlsx", sheet = "Commercial_2026")
```

Read in mangrove_data
```{r}
mangrove_data <- read_csv("data/mangrove_survey_raw.csv")
```

```{r}
mangrove_data <- read_csv("data/mangrove_survey_raw.csv",
  skip = 5,    # Skip the first 5 lines of field notes
  na = c(".", "NA", "9999", "ND", "blank"))  # Convert known text alts to true NA
```

Force a modern tibble to degrade into a legacy base R data frame structure
```{r}
benthic_cover_df <- as.data.frame(benthic_cover)
```

Print the old-style dataframe structure to view
```{r}
print(benthic_cover_df)
```

Compare with tibble alternative
```{r}
print(benthic_cover)
```

Install the data package 
```{r}
# install.packages("palmerpenguins")
```

Load the package data into active memory
```{r}
library(palmerpenguins)
data("penguins")
```

Examine the structure of the dataset - always do this when loading a new dataset!
```{r}
glimpse(penguins) # tidyverse version (from dplyr package)
str(penguins) # base R version
```

Generate an exploratory summary matrix
```{r}
summary(penguins)
```

Vertically slice specific morphometric variables by explicit name
```{r}
morphology_metrics <- select(penguins, species, bill_length_mm, bill_depth_mm, body_mass_g)
glimpse(morphology_metrics)
```

Retain a continuous block of attributes using the colon operator
```{r}
spatial_block <- select(penguins, species:island)
```

Discard logistics tracking attributes while preserving everything else using the minus sign
```{r}
clean_scientific_fields <- select(penguins, -year)
```

Isolate observations belonging to a single categorical target group
```{r}
adelie_cohort <- filter(penguins, species == "Adelie")
```

#### Sift out individuals using continuous numerical boundary thresholds
Preserves only large penguins whose mass exceeds 4500 grams
```{r}
heavy_penguins <- filter(penguins, body_mass_g > 4500)
```

#### Combine multiple conditional parameters across separate attributes
Preserves records matching Gentoo penguins sampled explicitly on Biscoe Island
```{r}
biscoe_gentoo <- filter(penguins, species == "Gentoo" & island == "Biscoe")
```

Sift records matching multiple targeting flags within an explicit set
```{r}
sub_islands <- filter(penguins, island %in% c("Dream", "Torgersen"))
```

Sort penguins by ascending body mass (Default setting: Smallest mass first)
```{r}
lightest_first <- arrange(penguins, body_mass_g)
```

Sort penguins in descending sequence using the desc() layout wrapper
```{r}
heaviest_first <- arrange(penguins, desc(body_mass_g))
```

Execute nested sorting criteria: Group by species, then sort by descending bill length
```{r}
stratified_morphology <- arrange(penguins, species, desc(bill_length_mm))
```

Pipe: best way
```{r}
penguins_final <- penguins |>
  mutate(bill_ratio = bill_length_mm / bill_depth_mm) |>
  filter(species == "Adelie")
```

Calculate a new morphological ratio in our environment
```{r}
penguin_ratios <- penguins  |> 
  mutate(body_mass_kg = body_mass_g / 1000,   # Convert grams to kilograms
         bill_ratio = bill_length_mm / bill_depth_mm  # Bill ratio
  )
```

View your newly engineered variables appended to the far-right columns
```{r}
glimpse(penguin_ratios)
```

Grouping our active memory penguins by species
```{r}
grouped_penguins <- group_by(penguins, species)
```

Notice that the table looks identical, but metadata notes 'Groups: species [3]'
```{r}
print(grouped_penguins)
```

Collapsing the buckets into explicit summary metrics
```{r}
species_mass_summary <- summarise(grouped_penguins,
                                  mean_mass_g = mean(body_mass_g)
)

print(species_mass_summary)
```

Overcoming the missing value trap using na.rm = TRUE
```{r}
biological_signal <- penguins %>%
  group_by(species, sex) %>%
  summarise(
    sample_size = n(),                                     # Count total individuals per category
    mean_mass_g = mean(body_mass_g, na.rm = TRUE),         # Calculate mean ignoring missing cells
    sd_mass_g   = sd(body_mass_g, na.rm = TRUE)            # Standard deviation calculation
  )

print(biological_signal)
```

#### Pipe directly from aggregation to plotting with error bars

Save the summary table
```{r}
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    sd_mass = sd(body_mass_g, na.rm = TRUE),
    n = n(),
    .groups = "drop"
  ) 
```

Print the summary table
```{r}
print(mass_compare_plot)
```

Plot from the saved summary object
```{r}
ggplot(mass_compare_plot, aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass, 
                    ymax = mean_mass + sd_mass), 
                width = 0.2) +
  facet_wrap(~island) +
  labs(title = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard deviation",
       y = "Mean Body Mass (g)",
       x = "Species") +
  theme_minimal()

mass_compare_plot

```

#### Change from standard deviation to standard error for mass

Save the summary table
```{r}
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    se_mass   = sd(body_mass_g, na.rm = TRUE) / sqrt(n()),
    n         = n(),
    .groups   = "drop"
  ) 
```

Print the summary table
```{r}
print(mass_compare_plot)
```

Plot from the saved summary object
```{r}
ggplot(mass_compare_plot, aes(x = species, y = mean_mass, colour = island)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_mass - se_mass,
                    ymax = mean_mass + se_mass),
                width = 0.2) +
  facet_wrap(~island) +
  labs(title    = "Mean Body Mass by Species and Island",
       subtitle = "Error bars represent standard error",
       y        = "Mean Body Mass (g)",
       x        = "Species") +
  theme_minimal()

mass_compare_plot
```

#### Make plots look more presentable

Standard deviation
```{r}
mass_compare_plot <- penguins |>
  group_by(species, island) |>
    summarise(
      mean_mass = mean(body_mass_g, na.rm = TRUE),
      sd_mass   = sd(body_mass_g, na.rm = TRUE),
      n         = n(),
      .groups   = "drop"
      ) 
```

Print the summary table
```{r}
print(mass_compare_plot)
```

Plot from the saved summary object
```{r}
ggplot(mass_compare_plot, aes(x = species, y = mean_mass, colour = island)) +
  geom_errorbar(aes(ymin = mean_mass - sd_mass,
                    ymax = mean_mass + sd_mass),
                width = 0.2, linewidth = 0.7, alpha = 0.7) +
  facet_wrap(~island) +
  geom_point(size = 4, alpha = 0.9) +
  scale_colour_manual(values = c(
    "Biscoe"    = "deeppink",
    "Dream"     = "blue",
    "Torgersen" = "aquamarine"
  )) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Mean Body Mass by Species and Island",
    subtitle = "Error bars represent ± 1 standard deviation",
    y        = "Mean Body Mass (g)",
    x        = NULL,
    colour   = "Island"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title       = element_text(face = "bold", margin = margin(b = 4)),
    plot.subtitle    = element_text(colour = "grey50", margin = margin(b = 12)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position  = "top",
    legend.title     = element_text(face = "bold"),
    axis.text        = element_text(colour = "grey30")
  )

mass_compare_plot
```


Standard deviation changed to standard error

Save the summary table
```{r}
mass_compare_plot <- penguins |>
  group_by(species, island) |>
  summarise(
    mean_mass = mean(body_mass_g, na.rm = TRUE),
    se_mass   = sd(body_mass_g, na.rm = TRUE) / sqrt(n()),
    n         = n(),
    .groups   = "drop"
  ) 
```

Print the summary table
```{r}
print(mass_compare_plot)
```

Plot from the saved summary object
```{r}
ggplot(mass_compare_plot, aes(x = species, y = mean_mass, colour = island)) +
  geom_errorbar(aes(ymin = mean_mass - se_mass,
                    ymax = mean_mass + se_mass),
                width = 0.3, linewidth = 0.9, alpha = 0.7) +
  facet_wrap(~island) +
  geom_point(size = 5, alpha = 0.9) +
  scale_colour_manual(values = c(
    "Biscoe"    = "deeppink",
    "Dream"     = "blue",
    "Torgersen" = "aquamarine"
  )) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title    = "Mean Body Mass by Species and Island",
    subtitle = "Error bars represent ± 1 standard error",
    y        = "Mean Body Mass (g)",
    x        = NULL,
    colour   = "Island"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title         = element_text(face = "bold", margin = margin(b = 4)),
    plot.subtitle      = element_text(colour = "grey50", margin = margin(b = 12)),
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.position    = "top",
    legend.title       = element_text(face = "bold"),
    axis.text          = element_text(colour = "grey30")
  )
mass_compare_plot
```


